---
title: "Perfume Collection Analysis"
author: "Julissa Dueñas"
output:
  html_document:
    df_print: paged
---

## Goal

As my perfume collection continued to grow, I wanted to take a more intentional, data-driven approach to curating it.

The goal of this project is to streamline my collection, identify a small set of signature scents for each season, and uncover the fragrance notes I naturally gravitate toward based on my wear history.

## Data

The two data sets I used for this project are `Master_List.csv` and `Daily_Log.csv`. 
`Master_List` contains my whole perfume collection. Each perfume has a unique ID, the brand of the fragrance, the name, the type (eau de toilette, body spray, etc.), all of its notes, which size I own, and the price.
`Daily_Log` is my wear history: each row records a date from August 28, 2025 and the ID of the fragrance I wore that day (matching `Master_List`).

```{r message=FALSE, warning=FALSE, include=FALSE}
#prepare libraries

library(DBI)
library(RSQLite)
library(tidyverse)
library(janitor)

#data
master <- read_csv("Master_List.csv")%>% clean_names() #my perfume collection
daily <- read_csv("Daily_Log.csv")%>% clean_names() #my wear history from 2025-08-28
```

### Data preparation
Using the date from 'Daily_Log' I derived what season it was. I then split each comma-separated fragrance note into seprate rows and assigned each one its own unique ID so wears can be tied to individual notes 

```{r echo=TRUE, message=FALSE, warning=FALSE, results='hide', fig.show='hide'}
daily <- daily %>%
  mutate(date = as.Date(date))
daily <- daily %>%
  mutate(date = as.Date(date),
         season = case_when(
           month(date) %in% c(12, 1, 2) ~ "Winter",
           month(date) %in% c(3, 4, 5) ~ "Spring",
           month(date) %in% c(6, 7, 8) ~ "Summer",
           month(date) %in% c(9, 10, 11) ~ "Fall"
         ))
#Split notes
notes_split <- master %>%
  select(id, notes) %>%
  separate_rows(notes, sep = ",") %>%
  mutate(notes = str_trim(notes),
         notes = tolower(notes)) %>%
  distinct()

notes_table <- notes_split %>%
  distinct(notes) %>%
  mutate(note_id = row_number()) %>%
  rename(note_name = notes)

perfume_notes <- notes_split %>%
  left_join(notes_table, by = c("notes" = "note_name")) %>%
  select(id, note_id)
```
### Database Setup
I built a local SQLite database to store the cleaned data and answer my burning questions with SQL. 'perfumes' hold my collection, 'daily_wear_log' holds wears by date and season, Fragrance notes are normalizes into 'note' and 'perfume_notes' provides a bridge so one fragrance can map to multiple ntoes.
```{r echo=TRUE, message=FALSE, warning=FALSE, results='hide', fig.show='hide'}
con <- dbConnect(RSQLite::SQLite(), "perfume.sqlite")

dbExecute(con, "
CREATE TABLE perfumes (
  id TEXT PRIMARY KEY,
  brand TEXT,
  name TEXT,
  fragrance_type TEXT,
  notes TEXT,
  size TEXT,
  price REAL
);
")
dbExecute(con, "
CREATE TABLE notes (
  note_id INTEGER PRIMARY KEY,
  note_name TEXT
);
")
dbExecute(con, "
CREATE TABLE perfume_notes (
  id TEXT,
  note_id INTEGER
);
")
dbExecute(con, "
CREATE TABLE daily_wear_log (
  date TEXT,
  id TEXT,
  season TEXT
);
")
dbWriteTable(con, "perfumes", master, append = TRUE)
dbWriteTable(con, "notes", notes_table, append = TRUE)
dbWriteTable(con, "perfume_notes", perfume_notes, append = TRUE)
dbWriteTable(con, "daily_wear_log", daily, append = TRUE)
```
## Analysis: Now the fun part!
### How big is my collection? How many full size perfumes do I have versus travel and sample sizes?

```{r echo=TRUE}
size_dist_raw <- dbGetQuery(con, "
SELECT size, COUNT(*) AS count
FROM perfumes
GROUP BY size
ORDER BY count DESC
")

total_perfumes <- sum(size_dist_raw$count)

size_dist <- size_dist_raw %>%
  mutate(
    total = total_perfumes,
    share_pct = round(100 * count / total, 1)
  )
```


```{r echo=FALSE}
levs <- unique(size_dist$size)
n_sizes <- length(levs)
pastel_fill <- rep(
  c("#F5C4D8", "#B8D4E8", "#C8E6C9", "#FFF3B0", "#E4C9F0", "#FFD4B8", "#B8EBE6", "#DDD4C8"),
  length.out = n_sizes
)
names(pastel_fill) <- levs
size_dist %>%
  mutate(
    pct = count / sum(count),
    slice_label = if_else(pct >= 0.035, as.character(count), ""),
    size_f = factor(size, levels = levs)
  ) %>%
  ggplot(aes(x = "", y = pct, fill = size_f)) +
  geom_col(width = 1, color = "white", linewidth = 0.5) +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = slice_label),
    position = position_stack(vjust = 0.5),
    size = 3.1,
    color = "gray15",
    fontface = "bold",
    lineheight = 0.95
  ) +
  scale_fill_manual(values = pastel_fill, name = "Size") +
  theme_void(base_size = 12) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = margin(12, 12, 12, 12),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  ) +
  labs(title = "Bottles by size")
```

In total, I have 192 perfume bottles (yikes). More than 51% of the collection is travel sized bottles.

### What are my top 10 perfumes and the notes I am most reaching for?
```{r}
top10 <- dbGetQuery(con, "
SELECT p.name, COUNT(*) as wears
FROM daily_wear_log d
JOIN perfumes p ON d.id = p.id
GROUP BY p.name
ORDER BY wears DESC
LIMIT 10
")
top_notes <- dbGetQuery(con, "
SELECT n.note_name, COUNT(*) as wears
FROM daily_wear_log d
JOIN perfume_notes pn ON d.id = pn.id
JOIN notes n ON pn.note_id = n.note_id
GROUP BY n.note_name
ORDER BY wears DESC
LIMIT 10
")
```

```{r fig-top10-notes, echo=FALSE, fig.height=5, fig.keep="all", fig.width=6.5, out.width="100%"}
pastel_palette <- c(
  "#F5C4D8", "#B8D4E8", "#C8E6C9", "#FFF3B0", "#E4C9F0", "#FFD4B8", "#B8EBE6", "#DDD4C8"
)

theme_bars <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13)
  )

# Top 10 fragrances — separate plot; tight x-axis; most-worn at top of y-axis
d_top10 <- top10 %>%
  mutate(label = name, n = wears) %>%
  arrange(desc(n)) %>%
  mutate(
    bar_fill = rep(pastel_palette, length.out = n())[row_number()],
    label = fct_reorder(label, n)
  )

p_top10 <- ggplot(d_top10, aes(x = n, y = label, fill = bar_fill)) +
  geom_col(width = 0.78, color = "white", linewidth = 0.45, show.legend = FALSE) +
  scale_fill_identity() +
  scale_x_continuous(
    limits = c(0, max(d_top10$n) * 1.12),
    expand = expansion(mult = c(0, 0.02))
  ) +
  theme_bars +
  labs(
    title = "Most-worn fragrances",
    x = "Logged wears",
    y = NULL
  )

print(p_top10)

# Top notes — separate plot; default x padding for potentially wider spread
d_notes <- top_notes %>%
  mutate(label = note_name, n = wears) %>%
  arrange(desc(n)) %>%
  mutate(
    bar_fill = rep(pastel_palette, length.out = n())[row_number()],
    label = fct_reorder(label, n)
  )

p_top_notes <- ggplot(d_notes, aes(x = n, y = label, fill = bar_fill)) +
  geom_col(width = 0.78, color = "white", linewidth = 0.45, show.legend = FALSE) +
  scale_fill_identity() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.08))) +
  theme_bars +
  labs(
    title = "Top Worn Notes",
    x = "Wear-attributed note mentions",
    y = NULL
  )

print(p_top_notes)
```

Being as I have 164 days of data, and I have worn a fragrance containing vanilla 107 times, I think it is more than fair to say vanilla earns the top spot

### What does this look like by season?
```{r}
top_by_season<-dbGetQuery(con, "
WITH ranked_perfumes AS (
  SELECT 
    d.season,
    p.brand,
    p.name,
    COUNT(*) AS wear_count
  FROM daily_wear_log d
  JOIN perfumes p ON d.id = p.id
  GROUP BY d.season, p.brand, p.name
)
SELECT *
FROM (
  SELECT *,
         ROW_NUMBER() OVER (
           PARTITION BY season
           ORDER BY wear_count DESC
         ) AS rank
  FROM ranked_perfumes
)
WHERE rank <= 5
ORDER BY season, rank;
")
top_notes_by_season <- dbGetQuery(con, "
WITH note_counts AS (
  SELECT 
    d.season,
    n.note_name,
    COUNT(*) AS note_count
  FROM daily_wear_log d
  JOIN perfume_notes pn ON d.id = pn.id
  JOIN notes n ON pn.note_id = n.note_id
  GROUP BY d.season, n.note_name
)
SELECT *
FROM (
  SELECT *,
         ROW_NUMBER() OVER (
           PARTITION BY season
           ORDER BY note_count DESC
         ) AS rank
  FROM note_counts
)
WHERE rank <= 10
ORDER BY season, rank;
")
```

```{r fig-seasonal-bars, echo=FALSE, fig.height=6, fig.keep="all", fig.width=8, out.width="100%"}
# Reuse the same pastel palette used earlier

theme_bars_season <- theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13)
  )

# Seasonal top fragrances
d_top_season <- top_by_season %>%
  mutate(label = paste(brand, name, sep = " - ")) %>%
  arrange(season, desc(wear_count)) %>%
  group_by(season) %>%
  mutate(
    bar_fill = rep(pastel_palette, length.out = n())[row_number()],
    label = fct_reorder(label, wear_count)
  ) %>%
  ungroup()

p_top_season <- ggplot(d_top_season, aes(x = wear_count, y = label, fill = bar_fill)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.45, show.legend = FALSE) +
  facet_wrap(~season, scales = "free_y", ncol = 2) +
  scale_fill_identity() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.08))) +
  theme_bars_season +
  labs(
    title = "Top Worn Perfumes by Season",
    x = "Logged wears",
    y = NULL
  )

print(p_top_season)

# Seasonal top notes
d_notes_season <- top_notes_by_season %>%
  arrange(season, desc(note_count)) %>%
  group_by(season) %>%
  mutate(
    bar_fill = rep(pastel_palette, length.out = n())[row_number()],
    note_name = fct_reorder(note_name, note_count)
  ) %>%
  ungroup()

p_notes_season <- ggplot(d_notes_season, aes(x = note_count, y = note_name, fill = bar_fill)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.45, show.legend = FALSE) +
  facet_wrap(~season, scales = "free_y", ncol = 2) +
  scale_fill_identity() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.08))) +
  theme_bars_season +
  labs(
    title = "Top Notes by Season",
    x = "Wear-attributed note mentions",
    y = NULL
  )

print(p_notes_season)
```
I don't have too much data from Summer months just yet, but already, I see some floral and fruity notes coming in for the Spring months. I am curious to see how the notes differ based on the seasons.

```{r eval=FALSE, include=FALSE}
never_worn <- dbGetQuery(con, "
SELECT p.name, p.size
FROM perfumes p
LEFT JOIN daily_wear_log d
ON p.id = d.id
WHERE d.id IS NULL
")

```

